Multiphase implementation on a parallel 2D/3D Lattice Boltzmann solver using GPGPUs

### 2017-08-17 Thesis/Draft.tex
spelling and ack

% Abstract and Keywords
\begin{abstract}
  This thesis aims to extend the functionalities of the in-house Lattice Boltzmann Method solver to include the processing of Multiphase flows, both in 2D and 3D. An in-depth analysis was performed on the previous solver in terms of software architecture, results validation and performance obtained. Then, a serial implementation of the Color Gradient model, derived from a Matlab implementation, was built using C code into the solver. After validating these results, the serial C code was parallelized using the CUDA platform. Once these results were validated, the process was then repeated for the 3D model. Finally, the implemented Color Gradient model was optimized to further improve the solver's performance. \par
  All of the implemented test cases were successful and the validation proved the model to be working, both in 2D and 3D. The overall run-times were reduced in 20 times for 2D and 176 times for 3D when adapting the code from the serial to the parallel version. The final parallel version's run-times were improved up to 18\% between the initial and final parallel implementation.  
\end{abstract}

@@

% The list of abbreviations can't be automatically generated so you need to populate it yourself
\begin{listofabbreviations}
	\abbrev{2D}{Two dimensions}
	\abbrev{3D}{Three dimensions}
    \abbrev{CUDA}{Compute Unified Device Architecture}
    \abbrev{CFD}{Computational Fluid Dynamics}
    \abbrev{GPGPU}{General-purpose computing on graphics processing units}
    \abbrev{GPU}{Graphics Processing Unit}
    \abbrev{LBM}{Lattice Boltzmann Method}
    \abbrev{BC}{Boundary condition}
    \abbrev{2DQ9}{Lattice speed model used for 2D}
    \abbrev{3DQ19}{Lattice speed model used for 3D}
    \abbrev{HPC}{High Performance Computing}
    \abbrev{BGKW}{Bhatnagar-Gross-Krook-Welander collision model}
    \abbrev{TRT}{Two Relaxation Time collision model}
    \abbrev{MRT}{Multiple Relaxation Time collision model}

@@

% Acknowledgements
\chapter{Acknowledgements}
The author would like to thank \dots

@@

Fluid dynamics is an ever-growing, ever-present part of the modern industry. Almost every field in the industry will eventually have to take fluid dynamics into account when designing and developing their product, from aircrafts, to combustion chambers and even in weather prediction systems. With modernization comes an increase in the amount of techniques available to process these issues, and one of them comes from the use of computers. Instead of spending a high amount of time, effort and, most importantly, money, in building and developing structures such as wind tunnels to study the design of an aircraft, for example, companies can take advantage of Computational Fluid Dynamics and simulate their needs using software. \par
This was traditionally done by solving classic Navier-Stokes equations that treat the fluid as a whole and analyze it's properties. Other ways could rely on the microscopic parts of the fluid and take each into particle into account. Another, great option would be to combine these 2 approaches into a median approach, which is what the Lattice Boltzmann method does, based on a statistical approach to solve the distribution functions associated with the fluid. This method also brings the advantage of it's locality nature. Each node's dependency on it's neighbors is very small, allowing for very few interactions between them. This makes the Lattice Boltzmann method an excellent candidate for parallel computations.\par
At the same time, the method's relative simplicity allows for the modeling of more complex and interesting flows, like Multiphase flows. These flows represent materials in different states, allowing for very interesting interactions like liquid/solid, gas/liquid or even liquid/liquid which will be the main focus of this thesis. These flows extend the applicability of the method to include capillarity problems, paper-making and even fluid management in space. \par
However, to perform a parallel implementation of the methods, one needs hardware on which to run the simulations. Nowadays, we can find multi-processing units in almost every computer, we can even find these in our smart-phones. Large scale simulations, like most used to simulate real-world problems require bigger computation abilities. High Performance Computing aims to enable these types of simulations, and they are usually used in centers specifically created to cover these needs. However, the clusters of computational units used are usually very expensive, but luckily there is a cheaper, viable alternative - GPUs. In 2007, NVIDIA released the CUDA platform, allowing users to take advantage of the processing power of their GPUs to perform General Purpose computing. GPUs offer the ability of using a very large amount of threads to parallelize the work by exploiting their architecture, at the cost of having to understand their complicated architecture and memory management system. \par

@@

In the Results and Discussion chapter we present the work done during this thesis and analyze the results obtained. We start by exposing the in-house solver, how it is built, it's performance and how the code works. Then we present how we developed the Multiphase model into the LBM solver. Afterwards, we discuss the various test cases employed in the model and how they were validated. We then analyze the memory usage of the solver and end the chapter with the performance that our model obtained, as well as the results of parallelization and optimization.\par

@@

\subsection{Multiphase flow}
Multiphase flows represent the simultaneous flow of materials in different states (phases). As such, multiphase flows have an enormous spectrum of representation and stand for interactions between gas/solid flows, liquid/solid flows or even liquid/liquid flows with different chemical properties \cite{multiphase}.\par
These simulations present challenging problems because of inherent difficulties during modelling and the importance of engineering applications. However, the Lattice Boltzmann method provides an alternative for these simulations due to it's relative simplicity when compared with traditional Navier-Stokes equation \cite{chenDoolenMP}.
\par These flows have several applications in the scientific and industrial community, ranging from pourous media fluid interactions \cite{pourous} to flows containing gas bubbles dispersed in liquids \cite{gasBubble}. Furthermore, almost every processing technology must take multiphase flows into account, including cavitating pumps, papermaking and many others \cite{multiphase}.
\par Several implementation strategies exist for Multiphase flows, however, for the purpose of this thesis, we will follow the Color Gradient model approach. Rothman and Keller \cite{rk} introduced the concept of 2 different fluids, blue and red, each with it's own distribution function. Later on, Gunstensen et al.  \cite{gunstensen} adapted the model proposed by Rothman and Keller into the Lattice Boltzmann method, whose main contribution was adding the perturbation operator, responsible for calculating the interface between both fluids. However, this version of the Color Gradient model was very computationally expensive and employs complicated programming techniques. D'Ortana et al. \cite{dortana} then improved the model by replacing the recoloring step strategy. Continuing the work, Latva-Kokko and Rothman modified this recoloring step and set the basis to extend the application of the model for the 2D and 3D speed models. Later on, the model was extended to the widely used 2DQ9 model by Reis and Phillips \cite{reisPhil}. This thesis will follow the implementation proposed by Leclaire et al. \cite{leclaire1, leclaire2, leclaire3}, where the equilibrium distribution function was enhanced.

@@

\section{Color Gradient Model}\label{sec:cgModel}
Models implementing multiphase flows for the Lattice Boltzmann method can generally be classified into several different categories. For the purpose of this thesis, we will focus on the Rothman-Keller methods, specifically on the model proposed by Reis and Phillips \cite{reisPhil}, modified using Latva-Kokko's recoloring operator, as proposed by Leclaire et al. \cite{leclaire3}. \par

@@

After computing the single-phase collision operator we need to add the result of the perturbation operator before passing it on to the recoloring sub-step. This is where we simulate the surface tension between the fluids and ensure that pressure difference is in equilibrium.\par
First, we must calculate the color gradient term, which is defined by

@@

\subsection{Recoloring operator}
The last step in the collision step is calculating the recoloring operator. This operator guarantees that the fluids remain immiscible and also controls the amount of a fluid sent to it's corresponding region. The operator can be defined as

@@

where $\beta$ is a free parameter between 0 and 1 that influences the thickness of the interface and $\cos \left(\phi_i\right)$ is the cosine of the angle between the color gradient $F$ and the direction $c$.

@@

The test cases for the Color Gradient model are based on a simple square for 2D or a cube for 3D problems. However, the boundaries are specific to each case. All of them use periodic boundaries on the East and West directions and most of them also use them for the North and South directions, the exception being for the Couette flow where the North boundary is an Inlet and the South boundary acts a solid wall.

@@

\section{Developing the Color Gradient model}
After making sure that the code we received was producing correct results, we could start implementing the Color Gradient (CG) model in the in-house LBM solver. Since there are two people working on this project, a proper separation of responsibilities  was vital for a successful adaption of the solver. To this end, Figueroa \cite{antonioThesis} was in charge of the physical aspects of the model whilst I was in charge of the software side. \par
We first decided to implement the CG model in a serial fashion so as to easily and quickly find potential errors in the code. Antonio would then develop a working script of the model in Matlab and, after validating it, I would take that script and patch it into the LBM solver. After making sure that the model was working in full integration with the LBM solver, I could then start parallelising the algorithms and finally validate the final, parallel results. We felt that this process, while extensive and time consuming, presented the best way of correctly implementing the model, since it was easier to track bugs, and so was repeated for the 3D model.
\subsection{Initial arguments}
The LBM solver depends on some initial parameters to initialize the required variables correctly. To do so, a SetUpData.ini file needs to be filled with the desired values. Figure \ref{fig:setupLBM} shows the initial arguments on which the LBM solver depends upon.

\begin{figure}[H]
	\centering
	\includegraphics[width=0.9\linewidth]{Resources/Images/setup.png}
	\caption{Setup file for the LBM solver}
	\label{fig:setupLBM}
\end{figure}
 However, the CG model requires several additional parameters. These parameters had to be added to the solver and can be viewed in Fig. \ref{fig:setupCG}.
\begin{figure}[H]
	\centering
	\includegraphics[width=0.9\linewidth]{Resources/Images/setupCG.png}
	\caption{Initial arguments for the Color Gradient model}
	\label{fig:setupCG}
\end{figure}

@@

As mentioned in Section \ref{sec:cgModel}, the Color Gradient model imposes some changes on the sub-steps of the LBM solver. The most noticeable of theses changes happen in the collision sub-step. First, we need to calculate the color gradient of each node. Since this color gradient is the result of a sum of interactions from the 9 directions of the speed model, we need a separate loop before being able to use this array in the computation of the perturbation operator. However, the color gradient calculation depends on the orientation of the node, that is whether the node is in the north, south, east, west or middle of the mesh. We can then calculate the appropriate collision frequency and afterwards compute the perturbation, collision and recoloring operators by looping between the 9 directions. For the most part of this sub-step, parallelization is implemented by following the example of the solver's previous collision sub-step. Each node will have it's own thread to perform calculations, and since each node does not depend on it's neighbors we do not have to worry about conflicted access to memory. However, each thread needs to know it's orientation for the color gradient calculation. Instead of having each thread reverse calculate it's index in an x,y format, a new array was created and initialized in the initialization section of the solver. This array has an element per node and it's value represents the orientation of the node in the following manner:

@@

The streaming step behaves like the original solver's streaming. We now have to stream not just one distribution function but both of them, as each fluid's changes are specific to it's own function. However, the 2D model was not behaving correctly when nodes near the boundary needed to be updated. This could be fixed by analyzing the mesh interpolation and how the stream array is built, which specifies whether a direction of the speed model should be updated or not. Since there was not enough time to perform this analysis, I used the same array as the one used in the color gradient to find out the orientation of node. Then I simply needed to specify how the streaming should work for each orientation. \par
The boundary step also behaves like the original solver's boundary update. However, the tests used to validate the Color Gradient model mostly use periodic boundaries, which did not exist in the 2D version of the previous solver. As such, these boundaries needed to be implemented and added to the 2D version of the solver. Furthermore, the boundaries also need an interpolation of the mesh in order for each thread to know what type of boundary that node should be updated with. Since the lattice generator software was out of scope for this thesis, I again used the same array built with the orientations for the color gradient model. This way, each thread could identify which boundary condition should be applied depending on the position of the node. \par

@@

\section{Validation}
The validation of the results was performed by Figueroa \cite{antonioThesis}. To do so, every test case was implemented according to their specific needs. Almost every case was based on the same mesh, a simple square or cube depending on the dimension. The only exception was the Rayleigh-Taylor instability where a rectangular mesh was used. Appendix \textcolor{red}{Ref appendix validation} shows that the model produces correct results for every test case. \par
The solver has the option of producing results in different formats, .dat, .vti and .csv. .VTI files were the addition of last year's authors and can be read using Paraview \cite{paraview}. The main advantage of these files is that they require significantly less memory than the other formats. Because of this, we are able to save time by transferring the results from Cranfield's HPC center faster and also by making post-processing analysis less time consuming. For this reason, we extended the creation of these files to use the new multiphase model arrays.
\section{Test cases}\label{sec:testcases}
The validation of the CG model depends not only on the analysis of the actual result of the simulation, but also in the type of test case used. Each test case had different requirements in the software since they depended on different arrays and each used their own type of verification. Because of this, I added the option to change between test cases as an argument in the setup file. During the execution of the solver, this options allows for a correct initialization of the densities and distribution functions, as well as adequately storing information when needed.
A total of 6 test cases were implemented for the 2D model and 4 for the 3D model.
\subsection{Steady bubble} 
\begin{figure}[H]
	\centering
	\begin{subfigure}{.49\textwidth}
		\includegraphics[width=\linewidth]{Resources/Images/steadyBubble.png}
		\caption{2D model}
	\end{subfigure}
	\begin{subfigure}{.49\textwidth}
		\includegraphics[width=\linewidth]{Resources/Images/steadyBubble3d.png}
		\caption{3D model}
	\end{subfigure} 
	\caption{Red fluid density for the Steady bubble case}
	\label{fig:steady}
\end{figure} 
This test aims to simulate a stationary bubble of an immiscible fluid inside another immiscible fluid. Although this test is the simplest physically, it is one the most computationally expensive tests to perform. To validate this test, we need to store the evolution of the surface tension over time. This entails calculating the pressure difference inside and outside the bubble, resulting in 4 additional reduce operations in every iteration. Reduce operations are very expensive in GPUs since these are operations that are dominated by memory access, requesting a heavy amount of memory access inside the GPU as well as when transferring information between the CPU and GPU. We are likely to see an increase in performance when running simulations with other, less computationally demanding test cases. 
\subsection{Deforming bubble}
\begin{figure}[H]
	\centering
	\begin{subfigure}{.49\textwidth}
		\includegraphics[width=\linewidth]{Resources/Images/square.png}
		\caption{2D model}
	\end{subfigure}
	\begin{subfigure}{.49\textwidth}
		\includegraphics[width=\linewidth]{Resources/Images/square3D.png}
		\caption{3D model}
	\end{subfigure} 
	\caption{Blue fluid density for the deforming bubble case}
	\label{fig:deforming}
\end{figure} 
This test simulates a fluid that starts as a deformed bubble, a square, and eventually converges into a steady, stable bubble. This test is much less computationally expensive. The only verification needed is the difference between the final bubble radius and the predicted radius, which is only done once, at the end of the simulation.

@@

\section{Memory usage}
The amount of memory used is a crucial factor when performing large scale simulations. Even more so if we want the results to mimic real-word scenarios. For this reason, I have performed an analysis on the amount of memory that the new Color Gradient model requires on the LBM solver from the GPU. Tables \ref{tab:2dmem} and \ref{tab:3dmem} show the amount of memory used by the 2D and 3D mesh containing 128 nodes in each direction.
\begin{table}[H]
	\centering
	\begin{tabular}{|l|l|l|}
		\hline
		\textbf{Variables}              & \textbf{Variable size}   & \textbf{Memory (MB)} \\ \hline
		\textbf{Initialization}         & $3 * m * n$              & 0.1875               \\ \hline
		\textbf{BC initialization}      & $7 * N_{Conn}$           & 0.0399475098         \\ \hline
		\textbf{Macroscopic values}     & $9 * m * n$              & 0.5625               \\ \hline
		\textbf{Boundary conditions}    & $16 * N_{BC}$            & 0.0302734375         \\ \hline
		\textbf{Distribution functions} & $6 * 9 * m * n$          & 3.375                \\ \hline
		\textbf{Residuals}              & $3 * 9 * m * n$          & 1.6875               \\ \hline
		\textbf{Temporary values}       & $5 * m * n$              & 0.3125               \\ \hline
		\textbf{Temporary values 2}     & $8 * m * n $        	   & 0.5                    \\ \hline
		\textbf{Sum}                    &                          & ~6.7         \\ \hline
	\end{tabular}	
	\caption{2D memory usage for Cavity 128 with Single precision}
	\label{tab:2dmem}
\end{table}

@@

In the 3D model, the memory usage is on a completely different scale. We now see that for the same mesh as in Table \ref{tab:2dmem}, extended to another direction, the used memory increases approximately 233 times. This makes the memory used by the 2D model almost insignificant when compared with the 3D version.\par
To further analyze what this increase in memory means, I have prepared Table \ref{tab:memCG} showing the different memory requirements for both single and double precision and comparing between a 128 and 256 sized mesh.

\begin{table}[H]
	\centering
	\begin{tabular}{|c|c|c|c|}
		\hline
		\textbf{Type}    & \textbf{Nodes} & \textbf{Single precision} & \textbf{Double precision} \\ \hline
		\textbf{2D\_128} & 16384          & 6.32                      & 11.7                      \\ \hline
		\textbf{2D\_256} & 65536          & 24                        & 46.57                     \\ \hline
		\textbf{3D\_128} & 2097152        & 1563                      & 3026                      \\ \hline
		\textbf{3D\_256} & 16777216       & 12336                     & 24054                     \\ \hline
	\end{tabular}
	\caption{Memory usage for the Color Gradient model in MB}
	\label{tab:memCG}
\end{table}

@@

\subsubsection{Normal distribution function v1.0}
This version of the code was the first parallel version using the Color Gradient model. In it, we specify the collision frequency to be used in the collision step in the initialization of the solver. Also, we calculate the perturbation operator for each distribution function, as well as the single phase collision step and recoloring operator.

>>>>

% Abstract and Keywords
\begin{abstract}
  This thesis aims to extend the functionalities of the in-house Lattice Boltzmann Method solver to include the processing of Multiphase flows, both in 2D and 3D. An in-depth analysis was performed on the previous solver in terms of software architecture, results validation and performance obtained. Then, a serial implementation of the Colour Gradient model, derived from a Matlab implementation, was built using C code into the solver. After validating these results, the serial C code was parallelized using the CUDA platform. Once these results were validated, the process was then repeated for the 3D model. Finally, the implemented Colour Gradient model was optimized to further improve the solver's performance. \par
  All of the implemented test cases were successful and the validation proved the model to be working, both in 2D and 3D. The overall run-times were reduced in 20 times for 2D and 176 times for 3D when adapting the code from the serial to the parallel version. The final parallel version's run-times were improved up to 18\% between the initial and final parallel implementation.  
\end{abstract}

@@

% The list of abbreviations can't be automatically generated so you need to populate it yourself
\begin{listofabbreviations}
	\abbrev{2D}{Two dimensions}
	\abbrev{3D}{Three dimensions}
    \abbrev{CUDA}{Compute Unified Device Architecture}
    \abbrev{CFD}{Computational Fluid Dynamics}
    \abbrev{GPGPU}{General-purpose computing on graphics processing units}
    \abbrev{GPU}{Graphics Processing Unit}
    \abbrev{LBM}{Lattice Boltzmann Method}
    \abbrev{BC}{Boundary condition}
    \abbrev{2DQ9}{Lattice speed model used for 2D}
    \abbrev{3DQ19}{Lattice speed model used for 3D}
    \abbrev{HPC}{High Performance Computing}
    \abbrev{BGKW}{Bhatnagar-Gross-Krook-Welander collision model}
    \abbrev{TRT}{Two Relaxation Time collision model}
    \abbrev{MRT}{Multiple Relaxation Time collision model}
    \abbrev{MB}{Mega Bytes}

@@

% Acknowledgements
\chapter{Acknowledgements}
This thesis would not have been possible without the help of several dear people to me. To start, I would like to thank my supervisor, Dr. Irene Moulitsas. Her help was invaluable for the conclusion of this thesis. She was able to both challenge me into diving into a unfamiliar topic and push me to keep going further. Our meetings always had a way of motivating me and making me believe everything was going to be ok, even when there were obstacles to overcome. \par
Secondly, I want to thank Antonio Figueroa for all his support. These past months would not have been possible if not for his expertise and enthusiasm. There was nothing a couple of hours on a Hangouts call could not resolve, especially when the program seemed unfix-able. I would not have been able to finish this project without his help. \par
This next part is in Portuguese since it makes little sense in English. Um obrigado fortíssimo aos meus pais por todo o seu apoio e dedicação ao longo destes 22 anos. Obrigado por serem um pilar constante, por toda a ajuda e carinho. Obrigado por insistirem sempre para que nunca desista, para que nunca comprometesse os meus ideais e acima de tudo, para que fosse feliz. \par
Um obrigado especial à Catarina que acreditou sempre em mim e que tornou a monotonia de escrever a tese num período muito mais alegre e saudável. Obrigado por todo o apoio e por toda a ajuda que me deste. Obrigado especialmente por seres capaz de me aturar, desde os momentos felizes até aos piores onde parecia que só via a tese à frente.\par
Por último, queria agradecer aos meus amigos mais próximos. Estes últimos tempos têm sido difíceis, mas nada que um café, uma sessão de DnD ou uma saída à baixa não resolvesse. Obrigado por tornarem a minha vida mais fácil e por toda a motivação e ajuda que me deram.

@@

Fluid dynamics is an ever-growing, ever-present part of the modern industry. Almost every field in the industry will eventually have to take fluid dynamics into account when designing and developing their product, from air-crafts, to combustion chambers and even in weather prediction systems. With modernization comes an increase in the amount of techniques available to process these issues, and one of them comes from the use of computers. Instead of spending a high amount of time, effort and, most importantly, money, in building and developing structures such as wind tunnels to study the design of an aircraft, for example, companies can take advantage of Computational Fluid Dynamics and simulate their needs using software. \par
This was traditionally done by solving classic Navier-Stokes equations that treat the fluid as a whole and analyse it's properties. Other ways could rely on the microscopic parts of the fluid and take each into particle into account. Another, great option would be to combine these 2 approaches into a median approach, which is what the Lattice Boltzmann method does, based on a statistical approach to solve the distribution functions associated with the fluid. This method also brings the advantage of it's locality nature. Each node's dependency on it's neighbours is very small, allowing for very few interactions between them. This makes the Lattice Boltzmann method an excellent candidate for parallel computations.\par
At the same time, the method's relative simplicity allows for the modelling of more complex and interesting flows, like Multiphase flows. These flows represent materials in different states, allowing for very interesting interactions like liquid/solid, gas/liquid or even liquid/liquid which will be the main focus of this thesis. These flows extend the applicability of the method to include capillarity problems, paper-making and even fluid management in space. \par
However, to perform a parallel implementation of the methods, one needs hardware on which to run the simulations. Nowadays, we can find multi-processing units in almost every computer, we can even find these in our smart-phones. Large scale simulations, like most used to simulate real-world problems require bigger computation abilities. High Performance Computing aims to enable these types of simulations, and they are usually used in centres specifically created to cover these needs. However, the clusters of computational units used are usually very expensive, but luckily there is a cheaper, viable alternative - GPUs. In 2007, NVIDIA released the CUDA platform, allowing users to take advantage of the processing power of their GPUs to perform General Purpose computing. GPUs offer the ability of using a very large amount of threads to parallelize the work by exploiting their architecture, at the cost of having to understand their complicated architecture and memory management system. \par

@@

In the Results and Discussion chapter we present the work done during this thesis and analyse the results obtained. We start by exposing the in-house solver, how it is built, it's performance and how the code works. Then we present how we developed the Multiphase model into the LBM solver. Afterwards, we discuss the various test cases employed in the model and how they were validated. We then analyse the memory usage of the solver and end the chapter with the performance that our model obtained, as well as the results of parallelization and optimization.\par

@@

\subsection{Multiphase flow}
Multiphase flows represent the simultaneous flow of materials in different states (phases). As such, multiphase flows have an enormous spectrum of representation and stand for interactions between gas/solid flows, liquid/solid flows or even liquid/liquid flows with different chemical properties \cite{multiphase}.\par
These simulations present challenging problems because of inherent difficulties during modelling and the importance of engineering applications. However, the Lattice Boltzmann method provides an alternative for these simulations due to it's relative simplicity when compared with traditional Navier-Stokes equation \cite{chenDoolenMP}.
\par These flows have several applications in the scientific and industrial community, ranging from porous media fluid interactions \cite{pourous} to flows containing gas bubbles dispersed in liquids \cite{gasBubble}. Furthermore, almost every processing technology must take multiphase flows into account, including cavitating pumps, papermaking and many others \cite{multiphase}.
\par Several implementation strategies exist for Multiphase flows, however, for the purpose of this thesis, we will follow the Colour Gradient model approach. Rothman and Keller \cite{rk} introduced the concept of 2 different fluids, blue and red, each with it's own distribution function. Later on, Gunstensen et al.  \cite{gunstensen} adapted the model proposed by Rothman and Keller into the Lattice Boltzmann method, whose main contribution was adding the perturbation operator, responsible for calculating the interface between both fluids. However, this version of the Colour Gradient model was very computationally expensive and employs complicated programming techniques. D'Ortana et al. \cite{dortana} then improved the model by replacing the recolouring step strategy. Continuing the work, Latva-Kokko and Rothman modified this recolouring step and set the basis to extend the application of the model for the 2D and 3D speed models. Later on, the model was extended to the widely used 2DQ9 model by Reis and Phillips \cite{reisPhil}. This thesis will follow the implementation proposed by Leclaire et al. \cite{leclaire1, leclaire2, leclaire3}, where the equilibrium distribution function was enhanced.

@@

\section{Colour Gradient Model}\label{sec:cgModel}
Models implementing multiphase flows for the Lattice Boltzmann method can generally be classified into several different categories. For the purpose of this thesis, we will focus on the Rothman-Keller methods, specifically on the model proposed by Reis and Phillips \cite{reisPhil}, modified using Latva-Kokko's recolouring operator, as proposed by Leclaire et al. \cite{leclaire3}. \par

@@

After computing the single-phase collision operator we need to add the result of the perturbation operator before passing it on to the recolouring sub-step. This is where we simulate the surface tension between the fluids and ensure that pressure difference is in equilibrium.\par
First, we must calculate the colour gradient term, which is defined by

@@

\subsection{Recolouring operator}
The last step in the collision step is calculating the recolouring operator. This operator guarantees that the fluids remain immiscible and also controls the amount of a fluid sent to it's corresponding region. The operator can be defined as

@@

where $\beta$ is a free parameter between 0 and 1 that influences the thickness of the interface and $\cos \left(\phi_i\right)$ is the cosine of the angle between the colour gradient $F$ and the direction $c$.

@@

The test cases for the Colour Gradient model are based on a simple square for 2D or a cube for 3D problems. However, the boundaries are specific to each case. All of them use periodic boundaries on the East and West directions and most of them also use them for the North and South directions, the exception being for the Couette flow where the North boundary is an Inlet and the South boundary acts a solid wall.

@@

\section{Developing the Colour Gradient model}
After making sure that the code we received was producing correct results, we could start implementing the Colour Gradient (CG) model in the in-house LBM solver. Since there are two people working on this project, a proper separation of responsibilities  was vital for a successful adaption of the solver. To this end, Figueroa \cite{antonioThesis} was in charge of the physical aspects of the model whilst I was in charge of the software side. \par
We first decided to implement the CG model in a serial fashion so as to easily and quickly find potential errors in the code. Antonio would then develop a working script of the model in Matlab and, after validating it, I would take that script and patch it into the LBM solver. After making sure that the model was working in full integration with the LBM solver, I could then start parallelising the algorithms and finally validate the final, parallel results. We felt that this process, while extensive and time consuming, presented the best way of correctly implementing the model, since it was easier to track bugs, and so was repeated for the 3D model.
\subsection{Initial arguments}
The LBM solver depends on some initial parameters to initialize the required variables correctly. To do so, a SetUpData.ini file needs to be filled with the desired values. Figure \ref{fig:setupLBM} shows the initial arguments on which the LBM solver depends upon.

\begin{figure}[H]
	\centering
	\includegraphics[width=0.9\linewidth]{Resources/Images/setup.png}
	\caption{Setup file for the LBM solver}
	\label{fig:setupLBM}
\end{figure}
 However, the CG model requires several additional parameters. These parameters had to be added to the solver and can be viewed in Fig. \ref{fig:setupCG}.
\begin{figure}[H]
	\centering
	\includegraphics[width=0.9\linewidth]{Resources/Images/setupCG.png}
	\caption{Initial arguments for the Colour Gradient model}
	\label{fig:setupCG}
\end{figure}

@@

As mentioned in Section \ref{sec:cgModel}, the Colour Gradient model imposes some changes on the sub-steps of the LBM solver. The most noticeable of theses changes happen in the collision sub-step. First, we need to calculate the colour gradient of each node. Since this colour gradient is the result of a sum of interactions from the 9 directions of the speed model, we need a separate loop before being able to use this array in the computation of the perturbation operator. However, the colour gradient calculation depends on the orientation of the node, that is whether the node is in the north, south, east, west or middle of the mesh. We can then calculate the appropriate collision frequency and afterwards compute the perturbation, collision and recolouring operators by looping between the 9 directions. For the most part of this sub-step, parallelization is implemented by following the example of the solver's previous collision sub-step. Each node will have it's own thread to perform calculations, and since each node does not depend on it's neighbours we do not have to worry about conflicted access to memory. However, each thread needs to know it's orientation for the colour gradient calculation. Instead of having each thread reverse calculate it's index in an x,y format, a new array was created and initialized in the initialization section of the solver. This array has an element per node and it's value represents the orientation of the node in the following manner:

@@

The streaming step behaves like the original solver's streaming. We now have to stream not just one distribution function but both of them, as each fluid's changes are specific to it's own function. However, the 2D model was not behaving correctly when nodes near the boundary needed to be updated. This could be fixed by analysing the mesh interpolation and how the stream array is built, which specifies whether a direction of the speed model should be updated or not. Since there was not enough time to perform this analysis, I used the same array as the one used in the colour gradient to find out the orientation of node. Then I simply needed to specify how the streaming should work for each orientation. \par
The boundary step also behaves like the original solver's boundary update. However, the tests used to validate the Colour Gradient model mostly use periodic boundaries, which did not exist in the 2D version of the previous solver. As such, these boundaries needed to be implemented and added to the 2D version of the solver. Furthermore, the boundaries also need an interpolation of the mesh in order for each thread to know what type of boundary that node should be updated with. Since the lattice generator software was out of scope for this thesis, I again used the same array built with the orientations for the colour gradient model. This way, each thread could identify which boundary condition should be applied depending on the position of the node. \par

@@

\section{Validation}
The validation of the results was performed by Figueroa \cite{antonioThesis}. To do so, every test case was implemented according to their specific needs. Almost every case was based on the same mesh, a simple square or cube depending on the dimension. The only exception was the Rayleigh-Taylor instability where a rectangular mesh was used. Appendix Section \ref{app:valid} shows that the model produces correct results for every test case. \par
The solver has the option of producing results in different formats, .dat, .vti and .csv. .VTI files were the addition of last year's authors and can be read using Paraview \cite{paraview}. The main advantage of these files is that they require significantly less memory than the other formats. Because of this, we are able to save time by transferring the results from Cranfield's HPC centre faster and also by making post-processing analysis less time consuming. For this reason, we extended the creation of these files to use the new multiphase model arrays.
\section{Test cases}\label{sec:testcases}
The validation of the CG model depends not only on the analysis of the actual result of the simulation, but also in the type of test case used. Each test case had different requirements in the software since they depended on different arrays and each used their own type of verification. Because of this, I added the option to change between test cases as an argument in the setup file. During the execution of the solver, this options allows for a correct initialization of the densities and distribution functions, as well as adequately storing information when needed.
A total of 6 test cases were implemented for the 2D model and 4 for the 3D model.
\subsection{Steady bubble} 
\begin{figure}[H]
	\centering
	\begin{subfigure}{.49\textwidth}
		\includegraphics[width=\linewidth]{Resources/Images/steadyBubble.png}
		\caption{2D model}
	\end{subfigure}
	\begin{subfigure}{.49\textwidth}
		\includegraphics[width=\linewidth]{Resources/Images/steadyBubble3d.png}
		\caption{3D model}
	\end{subfigure} 
	\caption{Red fluid density for the Steady bubble case}
	\label{fig:steady}
\end{figure} 
This test aims to simulate a stationary bubble of an immiscible fluid inside another immiscible fluid. Although this test is the simplest physically, it is one the most computationally expensive tests to perform. To validate this test, we need to store the evolution of the surface tension over time. This entails calculating the pressure difference inside and outside the bubble, resulting in 4 additional reduce operations in every iteration. Reduce operations are very expensive in GPUs since these are operations that are dominated by memory access, requesting a heavy amount of memory access inside the GPU as well as when transferring information between the CPU and GPU. We are likely to see an increase in performance when running simulations with other, less computationally demanding test cases. 
\subsection{Deforming bubble}
\begin{figure}[H]
	\centering
	\begin{subfigure}{.49\textwidth}
		\includegraphics[width=\linewidth]{Resources/Images/square.png}
		\caption{2D model}
	\end{subfigure}
	\begin{subfigure}{.49\textwidth}
		\includegraphics[width=\linewidth]{Resources/Images/square3D.png}
		\caption{3D model}
	\end{subfigure} 
	\caption{Blue fluid density for the deforming bubble case}
	\label{fig:deforming}
\end{figure} 
This test simulates a fluid that starts as a deformed bubble, a square, and eventually converges into a steady, stable bubble. This test is much less computationally expensive. The only verification needed is the difference between the final bubble radius and the predicted radius, which is only done once, at the end of the simulation.

@@

\section{Memory usage}
The amount of memory used is a crucial factor when performing large scale simulations. Even more so if we want the results to mimic real-word scenarios. For this reason, I have performed an analysis on the amount of memory that the new Colour Gradient model requires on the LBM solver from the GPU. Tables \ref{tab:2dmem} and \ref{tab:3dmem} show the amount of memory used by the 2D and 3D mesh containing 128 nodes in each direction.
\begin{table}[H]
	\centering
	\begin{tabular}{|l|l|l|}
		\hline
		\textbf{Variables}              & \textbf{Variable size}   & \textbf{Memory (MB)} \\ \hline
		\textbf{Initialization}         & $3 * m * n$              & 0.1875               \\ \hline
		\textbf{BC initialization}      & $7 * N_{Conn}$           & 0.0399475098         \\ \hline
		\textbf{Macroscopic values}     & $9 * m * n$              & 0.5625               \\ \hline
		\textbf{Boundary conditions}    & $16 * N_{BC}$            & 0.0302734375         \\ \hline
		\textbf{Distribution functions} & $6 * 9 * m * n$          & 3.375                \\ \hline
		\textbf{Residuals}              & $3 * 9 * m * n$          & 1.6875               \\ \hline
		\textbf{Temporary values}       & $5 * m * n$              & 0.3125               \\ \hline
		\textbf{Temporary values 2}     & $8 * m * n $        	   & 0.5                    \\ \hline
		\textbf{Sum}                    &                          & ~6.7         \\ \hline
	\end{tabular}	
	\caption{2D memory usage for Cavity 128 with Single precision}
	\label{tab:2dmem}
\end{table}

@@

In the 3D model, the memory usage is on a completely different scale. We now see that for the same mesh as in Table \ref{tab:2dmem}, extended to another direction, the used memory increases approximately 233 times. This makes the memory used by the 2D model almost insignificant when compared with the 3D version.\par
To further analyse what this increase in memory means, I have prepared Table \ref{tab:memCG} showing the different memory requirements for both single and double precision and comparing between a 128 and 256 sized mesh.

\begin{table}[H]
	\centering
	\begin{tabular}{|c|c|c|c|}
		\hline
		\textbf{Type}    & \textbf{Nodes} & \textbf{Single precision} & \textbf{Double precision} \\ \hline
		\textbf{2D\_128} & 16384          & 6.32                      & 11.7                      \\ \hline
		\textbf{2D\_256} & 65536          & 24                        & 46.57                     \\ \hline
		\textbf{3D\_128} & 2097152        & 1563                      & 3026                      \\ \hline
		\textbf{3D\_256} & 16777216       & 12336                     & 24054                     \\ \hline
	\end{tabular}
	\caption{Memory usage for the Colour Gradient model in MB}
	\label{tab:memCG}
\end{table}

@@

\subsubsection{Normal distribution function v1.0}
This version of the code was the first parallel version using the Colour Gradient model. In it, we specify the collision frequency to be used in the collision step in the initialization of the solver. Also, we calculate the perturbation operator for each distribution function, as well as the single phase collision step and recolouring operator.
